\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-23 NAR_DRAFT.tex
adding notes to self
Finish pass through Discussion

\rngcomment{Got here...}

\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
Early mutations are classified with low error, while later mutations are classified with less certainty.
We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
Thus, \batcave is best described as an algorithm trained on early mutations.

\bkmcomment{this needs more work}
We simulate tumors with three mutation profiles, and evaluate two real tumor data sets.
Among these performance is best with the most concentrated simulatin profile.
The intermediate and diffuse mutation profiles are designed to represent baseline or minimal profiles of Lung and Breast tumors respectively.
In fact, lung and breast tumors have heterogeneous mutation profiles with a wide range of levels of concentration.
The Acute Myeloid Leukemia we evaluated has a particularly diffuse profile, which is common for these tumors \cite{Alexandrov2019}.

These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
At the same time, there is no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
%We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
%Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.

We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.

The R implementation \texttt{batcaver} adds 1 second per 22,000 variants evaluated to a standard GATK best practices variant calling pipeline. 
The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.

We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications.
One size does not fit all with respect to the evidence threshold. 
Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of precision and recall.
We present \batcave, and the R package \texttt{batcaver} to provide the user with a more accurate sense of these trade-offs.

>>>>

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{XXX}. \rngcomment{Add citations, and sharpen discussion if necessary.}
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{TrackSigpaper}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

%\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
%In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
%Early mutations are classified with low error, while later mutations are classified with less certainty.
%We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
%Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
%Thus, \batcave is best described as an algorithm trained on early mutations.

We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively, which the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
But mutational profiles are highly heterogeneous, so that concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
The two real data sets we considered are among the very few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
They happen, however, to have relatively flat mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
Moreover, few true negative sites are validated, so specificity is impossible to calculate. \rngcomment{Double-check my logic here...}

%We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
%Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.

Our current implementation of \batcave is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
We chose to build \batcave off MuTect because MuTect is widely used, has state-of-the-art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Cibulskis2013,Griffith2015}.
But the mutational prior can be incorporated into almost any caller with an underlying probabilistic model.
For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a uniform probability of somatic mutation at each genomic site \cite{Kim2018}.
Replacing that uniform probability with a context-aware prior would require editing the Strelka2 source code, but is conceptually straightforward. \rngcomment{Is this correct, or does Strelka2 output enough for us to do it post-hoc?}

Our current R implementation, \texttt{batcaver}, is computationally inexpensive, adding 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
The majority of the computational costs is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
Since most callers are already walking the reference genome during the calling process, extracting the trinucleotide context simultaneously would virtually eliminate the computational cost of implementing a mutational context prior.

Our results show that adding a mutational context prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.

\rngcomment{Need to rearrange Discussion logically...}
